---
title: "Predicting Donor Behavior: Applying Machine Learning to Donor Classification"
author:
  name: Noah Ochital^[<no9857a@american.edu>]
  email: <no9857a@american.edu>
format:
    beamer:
        slide-level: 2
        institute: American University
        header-includes: \setbeamerfont{footnote}{size=\tiny}
fig-cap-location: bottom
tbl-cap-location: top
fig-pos: "H"
tbl-pos: "H"
execute:
  echo: FALSE
  output: FALSE
  warning: FALSE
  cache: TRUE
bibliography: predict_donor.bib
csl: american-political-science-association.csl
---

```{r}
library(tidyverse)
library(grid)
library(gridExtra)
library(ranger)
library(tidymodels)
library(naniar)
library(mice)
library(DescTools)
library(stargazer)
library(rpart)
library(rpart.plot)
library(BSDA)
library(rstatix)
library(vip)
library(here)
library(usmap)
library(kableExtra)
library(knitr)
library(xtable)
library(huxtable)

load(here("envs/presentation_env.RData"))
```

## Current State of Campaign Finance^[@glavin; @reynolds; @stratmann; @gordon; @baker; @opensecrets; @dwyre]

* Campaign Expenditures are rising:
    * 2000: $5 Billion.
    * 2020: $15 Billion.
* Campaign Expenditures correlate with electoral sucess.
* Individual Donors are a primary source of funding.
    * Provided about 60% of direct funding in 2020.

## Role for Predictive Models^[@dwyre; @walker2021; @nickerson2014]
* Donors are scarce:
    * 8.5% donated in 2020.
    * Less than 2% gave more than $200.
* Predictive models save campaign time and money.
* Prominent usage since 2008 but little academic research.

## Building Predictive Models
:::: {.columns}
::: {.column width="45%"}
* Classification requires informative features
* Tested 3 difference models:
    * Logistic regression.
    * Random forest.
    * Gradient boosteed trees.
* Main Questions:
   * Will these models be effective?
    * Which model will perform the best?
    * Which features are the most informative?
:::
::: {.column width="55%"}
```{r}
#| output: TRUE
#| fig-cap: "Example Classification Scenarios"
#| fig-width: 2
#| fig-height: 2

x <- rnorm(100)
y <- rnorm(100)
Donor <- factor(
  sample(c("No", "Yes"), 100, replace = TRUE, prob = c(1 - 0.085, 0.085)),
  levels = c("No", "Yes")
)
ex_data <- tibble(x, y, Donor)

ggplot(ex_data, aes(x = x, y = y, color = Donor)) +
  geom_jitter() +
  theme_void() +
  theme(
    legend.title = element_text(size = 10)
  )
```

:::
::::

## Donor Motivations^[@dwyre; @francia2003;@baker; @barber; @barber2016; @niebler; @ensley; @traag2016]
* Theorists have split donors into categories:
    * Investors.
    * Ideologues.
    * Intimates.
* Key empirical features:
    * Ideological fervor (Candidates and Voters).
    * Policy and partisan alignment.
    * Connections to other donors.


## Data Description^[@cces06; @cces07; @cces08; @cces10; @cces12; @cces14; @cces16; @cces17; @cces18; @cces19; @cces20; @cces22; @lones]
* Data from 2006-2008, 2012, 2014, 2016-2020, and 2022 CES surveys.
    * $n =$ `r nrow(original)`
    * `r round(mean(original$donate, na.rm = TRUE) * 100, 2) `% are donors.
* 27 features:
    * Categorical: 138 one-hot-coded.
    * Demographics.
    * Wealth.
    * Political Engagement.
    * Ideology.
* Training-Test Split
    * 80% training.
    * 20% test.

## Hypotheses
* $H_1$: At least one of the models will perform successfully.
* $H_2$: The more complex models will peform better relative to less complex models.
    * Donor behavior is complex.
* $H_3$: Political and civic engagement variables will have high predictive power:
    * Recognizing elected officials.
    * Political activities.
    * Voting intent.

## Missing Data^[@memon; @memon2022; @zhang; @vink; @austin; @mice; @lones]
:::: {.columns}
::: {.column width="45%"}
* Missing outcomes dropped
    * $n =$ `r nrow(train) + nrow(test)`
* Missingness mechanism assumption: Missing at Random (MAR)
* Multiple Imputation By Chained Equations (MICE):
    * 50 iterations.
    * Train and test imputed separately.
    * `mice` package in `R`.


:::
::: {.column width="55%"}
```{r}
#| output: 'asis'
#| tbl-cap: "Glimpse of Missing Variables"
#| label: tbl-gliv
gliv <- as.data.frame(rbind(head(missing_vars, 5), tail(missing_vars, 5)))[, c(
  1,
  2
)]
stargazer(
  gliv,
  summary = FALSE,
  digits = 0,
  header = FALSE,
  float = FALSE,
  font.size = "small",
  table.placement = "H"
)
```

:::
::::
## Class Weighting^[@kuhn2016a; @bakirarar]
* Significant class imbalance
    * `r round(mean(train$donate == "Yes")* 100)`% are donors.
    * `r round(mean(train$donate == "No")* 100)`% are not donors.
* Sample Based Class Weight (SBCW) method:
\begin{center}
\begin{equation*}
W_i = \frac{N}{k * n_i}
\end{equation*}
\end{center}
:::: {.columns}
::: {.column width="40%"}
\begin{center}
\begin{equation*}
W_{\text{donor}} = \frac{361323}{2 * 94548} = 1.9108
\end{equation*}
\end{center}
:::
::: {.column width="60%"}
\begin{center}
\begin{equation*}
W_{\text{non-donor}} = \frac{361323}{2 * 266775} = 0.6772
\end{equation*}
\end{center}
:::

## Models Used^[@james2013; @james2013a; @breiman_random; @friedman2001; @chen2016; @ke2017]

* Logistic Regression:
    * `glm()` from `R`
\begin{center}
$\hat{y}_i = \frac{1}{1 + \exp\left( -\symbf{x}_i^\text{T} \symbf{\beta} \right)}$
\end{center}
* Random Forest:
    * `ranger()` from `ranger` package in `R`.
\newline

\begin{center}
$\hat{y}_i = \frac{1}{T} \sum_{j=1}^N f_j(\symbf{x}_i) = \frac{\text{Tree}_1 + \text{Tree}_2 + \cdots +\text{Tree}_T}{T}$
\end{center}
* Gradient Boosted Trees:
    * `LightGBM` implementation in `R`
\newline
\begin{center}
$\hat{y}_i = f_0(\symbf{x}_i) + \sum_{t=1}^T f_t(\symbf{x}_i) = \text{Initial Guess} + \text{Tree}_1 + \text{Tree}_2 +\cdots +\text{Tree}_T$
\end{center}


## Hyperparameter Tuning^[@caret_book; @arnold; @probst; @raschka; @bergstra; @wu2019]
:::: {.columns}
::: {.column width="35%"}
* 10-fold CV on the training set
* Stage 1. Random hyperparameter search.
* Stage 2. 50 iterations of Bayesian Optimization.
:::
::: {.column width="65%"}
```{r}
#| output: 'asis'
#| tbl-cap: "Hyperparameter Ranges Tested"
params$Model <- c(params$Model[1:3], rep("LightGBM", 3))
kable(params, escape = FALSE)
```
:::
::::

## Test Set Performance
```{r}
#| output: 'asis'
#| tbl-cap: ""
names(metrics) <- c("Metric", "Logit", "Gradient Boosted", "Random Forest")
stargazer(
  as.data.frame(slice(metrics, 1:5)),
  summary = FALSE,
  digits = 4,
  header = FALSE,
  float = FALSE,
  font.size = "small",
  notes = "0.01 corresponds to 900 (Acc.), 237 (Rec.), and 326 (Prec.) People.",
  notes.align = "l",
  column.sep.width = "0.5pt"
)
```

## Model Comparisons^[@raschka]
* $\alpha = 0.05$
* Cochran's Q Test:
    * $H_0: ACC_1 = ACC_2 = ACC_3$
    * $H_a: \exists \ i \neq j \text{ such that } ACC_i \neq ACC_j \text{, for } i,j \in \{1,2,3\}$
    * $p=$`r options(scipen = 1, digits = 2); CochranQ$p`
* Pairwise McNemar:
    * $H_0: ACC_i = ACC_j$
    * $H_a: ACC_i \neq ACC_j$
    * Logit and boosted trees: `r McNemar$p[1]`
    * Random forest and boosted trees: `r McNemar$p[2]`
    * Logit and random forest: `r McNemar$p[3]`

## Select Variable Contributions: Permutation Importance
```{r}
#| output: 'asis'
#| tbl-cap: ""
var_imp_final <- slice(var_imp2, c(1:8, 11, 13, 16, 17, 20:21, 27)) |>
  as.data.frame()
names(var_imp_final) <- c(
  "Variable",
  "Logit",
  "Random Forest",
  "Gradient Boosted"
)
stargazer(
  var_imp_final,
  summary = FALSE,
  header = FALSE,
  digits = 4,
  font.size = "footnotesize",
  float = FALSE
)
```

## Revist Research Questions and Hypotheses
* Q1: Were these models effective?
    * Potentially, but without a proper comparison its difficult to conclude anything.
* Q2: Which model was the best?
     * F1 and Precision: Random Forest.
     * Recall and ROC-AUC: Gradient Boosted.
* Q3: Which features were the most important to predictive power?
     * Political Activities.
     * Model Dependent.
        * Ideology, Party ID, Voting Intent


## Limitations
* Self reporting, missingness, other biases from survey.
* Limited nuance in "Donor" variable.
    * Contribution size.
    * Contribution frequency.
    * Contribution recipient.
* Aggregating across election years.
    * Changing political and campaign finance landscape.

## Future Research and Next Steps:
* More comprehensive data:
    * Non-free sources.
    * State or election specific.
    * Specifically engineered surveys.
* Utilizing LLMs
    * LLMs can view the data in the raw form, and have deeper embeddings for qualitative data.
    * Vast array of Pre-trained models.
    * Low Rank Adaptation to retrain.


## Thank You

* Email: <no9857a@american.edu>, <noahochital@icloud.com>
* Github: Noch05

## References {allowframebreaks=1}
\tiny
:::{#refs}
:::
